Memes: an R interface to the MEME Suite

### 2020-10-19 manuscript.Rmd
add some edits from DJM

Biopolymers, such as DNA and protein, perform varying functions based on their
primary sequence. Short, repeated sequences, or "motifs" represent functional
units within biopolymers which can act as interaction surfaces, create
structure, or have enzymatic activity. Identification of shared motifs across
multiple sequences can provide evidence for shared function, thus the ability to
identify, classify, and compare motifs represents a key step in the analysis of
biological sequences. Although tools implementing these features exist within
Bioconductor, they often rely on proprietary data types with poor
interoperability with other tools. Furthermore, several of these packages will
be deprecated in the next Bioconductor release, leaving a gap in functionality
with no clear alternatives.

The MEME Suite is a series of commandline tools which provide a variety of
features for motif analysis including *de-novo* motif discovery, enrichment
testing, comparison, and matching. Here we present `memes`, a motif analysis
package which provides a seamless interface to MEME Suite tools within R.

@@

In addition to wrapping MEME Suite utilities, `memes` extends the functionality
of `universalmotif` to provide base R and `tidyverse` compatible syntax for
manipulating and analyzing motif metadata. These features allow motif databases
to integrate external data to aid analysis, such as filtering a set of
transcription factor motifs based on gene expression. `memes` also provides
simple utilities for data visualization.

`memes` is designed for maximum flexibility and ease of use to allow rapid
iteration during analysis. Here we present several examples of how `memes`
allows novel analyses of ChIP and FAIRE-seq data by facilitating seamless
interoperability between MEME Suite tools and the broader R landscape.

# Abstract Draft
Biopolymers, such as DNA and protein, perform varying functions based on their
primary sequence. Short, repeated sequences, or "motifs" represent functional
units within biopolymers which can act as interaction surfaces, create
structure, or have enzymatic activity. Identification of shared motifs across
multiple sequences can provide evidence for shared function, thus the ability to
identify, classify, and compare motifs represents a key step in the analysis of
biological sequences. Although tools that implement these features have
historically existed in the Bioconductor landscape, they often rely on
proprietary data types with poor interoperability with other tools. Furthermore,
several of these packages will be deprecated in the next Bioconductor release,
leaving a gap in functionality with no clear alternatives.

@@

Biopolymers, such as DNA and protein, perform varying functions based on their
primary sequence. Short, repeated sequences, or "motifs" represent functional
units within biopolymers which can act as interaction surfaces, create
structure, or have enzymatic activity. Identification of shared motifs across
multiple sequences can provide evidence for shared function, thus the ability to
identify, classify, and compare motifs represents a key step in the analysis of
biological sequences. Although tools that implement these features have
historically existed in the Bioconductor landscape, they often rely on
proprietary data types with poor interoperability with other tools. Furthermore,
several of these packages will be deprecated in the next Bioconductor release,
leaving a gap in functionality with no clear alternatives.

The MEME Suite is a series of commandline and web-based tools which provide a variety
of features for motif analysis such as *de-novo* motif discovery, enrichment
testing, comparison, and matching \cite{meme-suite}. The MEME Suite is a widely
accepted set of tools with over 22,000 citations to date, and over 30,000 unique
users of the webserver implementation annually \cite{meme-usage-report-page,
google-scholar-meme}.

>>>>

Biopolymers, such as DNA and protein molecules, perform varying functions based on their
primary sequence. Short, repeated sequences, or "motifs" represent functional
units within biopolymers that can act as interaction surfaces, create
structure, or have enzymatic activity. Identification of shared motifs across
multiple sequences can provide evidence for shared function. Thus, the ability to
identify, classify, and compare motifs represents a key step in the analysis of
biological sequences. **TODO: Transition sentence RE: R/motifs** Although tools
implementing these features exist within Bioconductor, they often rely on
proprietary data types with poor interoperability with other tools. **TODO: remove this sentence?** Furthermore,
several of these packages will be deprecated in the next Bioconductor release,
leaving a gap in functionality with no clear alternatives.

The MEME Suite is a series of commandline tools that provide a variety of
features for motif analysis including *de-novo* motif discovery, enrichment
testing, comparison, and matching. Here we present `memes`, a motif analysis
package that provides a seamless interface to MEME Suite tools within R.

@@

In addition to wrapping MEME Suite utilities, `memes` extends the functionality
of `universalmotif` to provide base R and `tidyverse` compatible syntax for
manipulating and analyzing motif metadata. These features allow motif databases
to integrate external data to aid analysis, such as filtering a set of
transcription factor motifs based on gene expression. Lastly, `memes` also provides
simple utilities for data visualization.

`memes` is designed for maximum flexibility and ease of use to allow rapid
iteration during analysis. Here we present several examples of how `memes`
allows novel analyses of ChIP-seq and FAIRE-seq data by facilitating seamless
interoperability between MEME Suite tools and the broader R landscape.

# Abstract Draft (OLD: delete??)
Biopolymers, such as DNA and protein, perform varying functions based on their
primary sequence. Short, repeated sequences, or "motifs" represent functional
units within biopolymers which can act as interaction surfaces, create
structure, or have enzymatic activity. Identification of shared motifs across
multiple sequences can provide evidence for shared function, thus the ability to
identify, classify, and compare motifs represents a key step in the analysis of
biological sequences. Although tools that implement these features have
historically existed in the Bioconductor landscape, they often rely on
proprietary data types with poor interoperability with other tools. Furthermore,
several of these packages will be deprecated in the next Bioconductor release,
leaving a gap in functionality with no clear alternatives.

@@

Biopolymers, such as DNA and protein molecules, perform varying functions based
on the composition of their primary sequence. Short, repeated sequences, or
"motifs" represent functional units within biopolymers which can act as
interaction surfaces, create structure, or have enzymatic activity.
Identification of shared motifs across multiple sequences can provide evidence
for shared function, thus the ability to identify, classify, and compare motifs
represents a key step in the analysis of biological sequences. Although tools
that implement these features have historically existed in the Bioconductor
landscape, they often rely on proprietary data types with poor interoperability
with other tools. Furthermore, several of these packages will be deprecated in
the next Bioconductor release, leaving a gap in functionality with no clear
alternatives.

The MEME Suite is a series of commandline and web-based tools that provide a variety
of features for motif analysis such as *de-novo* motif discovery, enrichment
testing, comparison, and matching \cite{meme-suite}. The MEME Suite is a widely
utilized set of tools, with over 22,000 citations to date, and greater than 30,000 unique
users of the webserver implementation annually \cite{meme-usage-report-page,
google-scholar-meme}.
